Towards Proof-Reconstruction of Problems in Classical Propositional Logic

### 2017-09-08 paper/sections/abstract.tex
[ paper ] section 3 introduction added.

\begin{abstract}
In this paper, we describe our approach to reconstruct and verify in
\Agda, a  proof assistant for type theory, solutions obtained by \Metis, an
automatic theorem prover, to problems in classical propositional
logic. The translation method consists of justifying step-by-step
each deduction found in the prover's solution. To carry out this
plan, we emulate the inference rules of the prover, providing a
version of them in the proof assistant. Finally, we parse the output
of the prover, build and analize the respective tree associated with the
proof, for ending up with an \Agda code of the proof for type-checking.

>>>>

\begin{abstract}
In this paper, we describe our approach to reconstruct and verify in
\Agda, a  proof assistant for type theory, solutions obtained by
\Metis, an automatic theorem prover, to problems in classical 
propositional logic. The translation method consists of justifying 
step-by-step each deduction found in the prover's solution. To carry 
out this plan, we emulate the inference rules of the prover, 
providing a version of them in the proof assistant. Finally, we 
parse the output of the prover, build and analyze the respective 
tree associated with the proof, for ending up with an \Agda code of 
the proof for type-checking.

### 2017-09-08 paper/sections/conclusions.tex
[ paper ] section 3 introduction added.

Besides the logic and its complexity, the proof reconstruction task is
far from trivial. In this document, we attempt to reconstruct proofs
of propositional logic without equality deliverd by \Metis successfully.

From our experience building a proof reconstruction tool, depending on
the systems involved in thetranslation, some difficulties can arise.
We listed some of them in section \ref{sec:metis-language-and-proofs}
and in the section \ref{sec:proof-reconstruction}.

For instance, just showing equality between Normal Forms for the
same formula in the output of the \canonicalize inference and a
common implementation for this kind of conversion; we observed that
\Metis performs reordering for each inner expression mainly in the
presence of negative literals. Another aspect to take into account
is the \Metis' treatment for conjunctions and disjunction;
these connectives were treated as right associative and binary
operators in the \TSTP format but that assumptons is not a standard.
We had to update the \TSTP parser module introduced in
\cite{Gomez-Londono2015} to include features like these.

However, one of our biggest obstacles
\improvement{Remove the use of the \emph{biggest obstacules}.}
was dealing with
the Equality between formulas since, in our approach, we only work
with the syntactic treatment of the logic without considering the
semantics.
One of the main reasons for such a restriction relay on our plans to
support further first-order logic, where a semantic treatment is not
viable.\unsure{Please confirm this.}

>>>>

In this document, we attempted to reconstruct proofs
of propositional logic without equality deliver by \Metis successfully.

From our experience building a proof reconstruction tool, depending 
on the systems involved in the translation, some difficulties can 
arise. We listed some of them in section
\ref{sec:metis-language-and-proofs} and in the section
\ref{sec:proof-reconstruction}.

For instance, just showing equality between Normal Forms for the
same formula in the output of the \canonicalize inference and a
common implementation for this kind of conversion; we observed that
\Metis performs reordering for each inner expression mainly in the
presence of negative literals. Another aspect to take into account
is the \Metis' treatment for conjunctions and disjunction;
these connectives were treated as right associative and binary
operators in the \TSTP format but that assumptions is not a standard.
We had to update the \TSTP parser module introduced in
\cite{Gomez-Londono2015} to include features like these.

However, one of our biggest obstacles
\improvement{Remove the use of the \emph{biggest obstacles}.}
was dealing with
the Equality between formulas since, in our approach, we only work
with the syntactic treatment of the logic without considering the
semantics.
One of the main reasons for such a restriction relay on our plans to
support further first-order logic, where a semantic treatment is not
viable.\unsure{Please confirm this.}

### 2017-09-08 paper/sections/proof-reconstruction.tex
[ paper ] section 3 introduction added.

\begin{figure}
\begin{tikzpicture}
\node[text width=2cm, align=center](problem) at (0,0)
  {1. CPL \\ Problem.};

\node[right = 1.2cm of problem, text width=2cm, align=center]
  (tptp){2. \TPTP \\ Problem.};

\node[right= 1.2cm of tptp, text width=2cm, align=center]
   (metis) {3. \Metis \\ Prover.};

\node[right= 1.2cm of metis, text width=2cm, align=center, inner sep=10pt]
   (tstp) {4. \TSTP \\ Derivation.};

@@

\node[left = 1.2cm of athena, text width=2cm, align=center]
   (agdafile) {6. \Agda \\ Proof.};

\node[left = 1.2cm of agdafile, text width=2cm, align=center]
   (agda) {7. Proof Checker.};

@@

\node[below = 0.5cm of verified, text width=2cm, align=center]
   (failure) {8.2 Error.};

% node[below] {send to}
\draw[->, thick] (problem) to
  % node[below] {\tiny encoding}
  (tptp);
\draw[->, thick] (tptp) to
  % node[below] {\tiny }
  (metis);
\draw[->, thick] (metis) to
  % node[below] {\tiny replies on}
  (tstp);
\draw[->, thick] (tstp) to
  % node[right] {\tiny parsing}
  (athena);
\draw[->, thick] (athena) to
  % node[below] {\tiny traslation}
  (agdafile);
\draw[->, thick] (agdafile) to
  % node[below] {\tiny type-checking}
  (agda);
\draw[->, thick] (agda) to (verified);
\draw[->, thick, gray] (agda) to (failure);
\end{tikzpicture}
\caption{Proof reconstruction overview.}
\label{fig:proof-reconstruction-overview}
\end{figure}

>>>>

\begin{figure}
\centering

\begin{tikzpicture}[scale=0.9]
\node[text width=2cm, align=center](problem) at (0,0)
  {1. CPL \\ Problem.};

\node[right = 1.1cm of problem, text width=2cm, align=center]
  (tptp){2. \TPTP \\ Problem.};

\node[right= 1.1cm of tptp, text width=2cm, align=center]
   (metis) {3. \Metis \\ Prover.};

\node[right= 1.1cm of metis, text width=2cm, align=center, inner sep=10pt]
   (tstp) {4. \TSTP \\ Derivation.};

@@

\node[left = 1.1cm of athena, text width=2cm, align=center]
   (agdafile) {6. \Agda \\ Proof.};

\node[left = 1.1cm of agdafile, text width=2cm, align=center]
   (agda) {7. Proof Checking.};

@@

\node[below = 0.5cm of verified, text width=2cm, align=center]
   (failure) {8.2 Failure.};

% node[below] {send to}
\draw[->, thick] (problem) to
  % node[below] {\tiny encoding}
  (tptp);
\draw[->, thick] (tptp) to
  % node[below] {\tiny }
  (metis);
\draw[->, thick] (metis) to
  % node[below] {\tiny replies on}
  (tstp);
\draw[->, thick] (tstp) to
  % node[right] {\tiny parsing}
  (athena);
\draw[->, thick] (athena) to
  % node[below] {\tiny traslation}
  (agdafile);
\draw[->, thick] (agdafile) to
  % node[below] {\tiny type-checking}
  (agda);
\draw[->, thick] (agda) to (verified);
\draw[->, thick, gray] (agda) to (failure);
\end{tikzpicture}
\caption{Proof reconstruction overview.}
\label{fig:proof-reconstruction-workflow}
\end{figure}

The proof reconstruction approach proposed here consists of a series 
of steps similar to the workflow presented by Sultana in
\cite{sultana2015}. This process is a translation from a source 
system to a target system. In our case, the system of origin, the 
automatic theorem prover, is \Metis; the target system is the proof 
assistant, \Agda. We choose \Agda, but another proof assistant with 
support of type theory and features similar like \Agda is valid; our 
results do not depend on the proof assistant but the support of inductive data types, and termination checking, and type-checking.
\improvement{Make this statement stronger and clear.}

Following the diagram in Fig.
\ref{fig:proof-reconstruction-workflow}, the process can begin with 
a problem in \CPL that we encode in \TPTP format to be sent as the 
input of the \Metis prover.
The proof-search algorithms of \Metis in step No. 3 perform a 
search for a solution; if they success, \Metis 
replies a derivation in \TSTP format of the proof, else \Metis may 
no stop running forever or replies the problem is counter feasible.

With such a derivation, we process it with our \Haskell translator 
tool, \Athena. \Athena parses the \TSTP format, analyze the 
derivation and generates a representation of the natural deduction 
proof using a version of a tree data structure (see the properties 
of this tree in subsection \ref{ssec:metis-proofs}). As a result,
\Athena generates an \Agda file with such a proof. We have included 
in this file, imports for \Agda libraries that give the support for 
the logic~\cite{AgdaProp}, and versions of the \Metis' inference 
rules~\cite{AgdaMetis}.

Finally, we use the proof assistant in step No. 7 to type-check the 
proof and verify its correctness. We have only one answer from two 
possible answers: the type-checking success (i.e., the proof is 
valid) or not. If the type-check fails, one of the three actors are 
responsible: \Metis, \Athena, or \Agda. We are responsible for 
errors committed in the translation from \TSTP to \Agda; lack of the 
coverage of the \Metis' inference rules, or bad assumptions about 
them.

In the rest of this section, we provide a formal description using 
type theory to build definitions and theorems for the functions 
necessary to emulate the \Metis' inference rules.
